### 2022-09-08 doc/src/CSRs.tex
Add a section about AIA bits in the state-enable CSRs

>>>>

%-----------------------------------------------------------------------
\section{Access control by the state-enable CSRs}

If extension Smstateen is implemented together with the
Advanced Interrupt Architecture (AIA), three bits of state-enable
register \z{mstateen0} control access to AIA-added state
from privilege modes less privileged than \mbox{M-mode}:
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect}, \z{sireg}, \z{vsiselect}, and \z{vsireg} \\
bit 59 & all other state added by the AIA
          and not controlled by bits 60 and 58 \\
bit 58 & all IMSIC state, including CSRs \z{stopei} and \z{vstopei} \\
\end{displayLinesTable}

If one of these bits is zero in \z{mstateen0}, an attempt to access
the corresponding state from a privilege mode less privileged
than \mbox{M-mode} results in an illegal instruction trap.
As always, the state-enable CSRs do not affect the accessibility
of any state when in \mbox{M-mode}, only in less privileged modes.
For more explanation, see the documentation for extension Smstateen.

Bit~59 controls access to AIA CSRs \z{siph}, \z{sieh}, \z{stopi},
\z{hidelegh}, \z{hvien}/\z{hvienh}, \z{hviph}, \z{hvictl},
\z{hviprio1}/\z{hviprio1h}, \z{hviprio2}/\z{hviprio2h}, \z{vsiph},
\z{vsieh}, and \z{vstopi}, as well as to the supervisor-level interrupt
priorities accessed through \z{siselect} + \z{sireg}
(the \z{iprio} array of Section~\ref{sec:intrPrios-S}).

Bit~58 is implemented in \z{mstateen0} only if the hart has an IMSIC.
If the hypervisor extension is also implemented, this bit does
not affect the behavior or accessibility of hypervisor CSRs
\z{hgeip} and \z{hgeie}, or field VGEIN of \z{hstatus}.
In particular, guest external interrupts from an IMSIC
continue to be visible to \mbox{HS-mode} in \z{hgeip}
even when bit~58 of \z{mstateen0} is zero.

\begin{commentary}
An earlier, pre-ratification draft of Smstateen said that when
bit~58 of \z{mstateen0} is zero, registers \z{hgeip} and \z{hgeie}
and field VGEIN of \z{hstatus} are all read-only zeros.
That effect is no longer correct.
\end{commentary}

If the hypervisor extension is implemented, the same three bits are
defined also in \z{hstateen0} but concern only the state potentially
accessible to a virtual machine executing in privilege modes VS and VU:
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect} and \z{sireg}
          (really \z{vsiselect} and \z{vsireg}) \\
bit 59 & CSRs \z{siph} and \z{sieh} (RV32 only) and \z{stopi}
          (really \z{vsiph}, \z{vsieh}, and \z{vstopi}) \\
bit 58 & all state of IMSIC guest interrupt files,
          including CSR \z{stopei} (really \z{vstopei}) \\
\end{displayLinesTable}

If one of these bits is zero in \z{hstateen0}, and the same bit is one
in \z{mstateen0}, then an attempt to access the corresponding state
from \mbox{VS-mode} raises a virtual instruction exception.

Bit~58 is implemented in \z{hstateen0} only if the hart has an IMSIC.
Furthermore, even with an IMSIC, bit~58 may (or may not) be read-only
zero in \z{hstateen0} if the IMSIC has no \emph{guest interrupt files}
for guest external interrupts (Chapter~\ref{ch:IMSIC}).
When this bit is zero, a virtual machine is prevented from
accessing the hart's IMSIC the same as when \z{hstatus}.VGEIN =~0.

Extension Ssstateen is defined as
the supervisor-level view of Smstateen.
Ssstateen therefore incorporates the bits defined above
for \z{hstateen0} but not those for \z{mstateen0},
since machine-level CSRs are not visible to supervisor level.
